Context-dependent consequences of lagged effects in demographic models

### 2025-08-27 docs/ipm_comparison_ms.qmd
ready for review by ES

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
        <!-- \usepackage[utf8]{inputenc} -->
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
```

```{r}
#load targets
withr::with_dir(here(), {
  tar_load(c(
    aic_tbl_df,
    lambda_table,
    fig_pop_states
  ))
})
```

@@

\pagebreak

@@

@@

There is increasing evidence that an organism's current likelihood of growth, survival, or reproduction can be strongly influenced by previous environmental conditions. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckerman_population_2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @criley_year_1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one several months later whose timing coincides with birth or flowering. The delay or intensity of lagged effects can also depend on local ecological conditions, with individuals in some habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.\

Because Lagged Effects are often directly linked to reproduction and survival, it is thought they could have major consequences for population dynamics [@beckerman_population_2002]. Although there is emerging evidence that this is indeed the case [e.g., @molowny-horas_changes_2017; @tenhumberg_time-lagged_2018; @williams_life_2015], broader efforts to test this hypothesis have been hampered by two primary factors: First, detecting lagged effects requires long-term data on both the putative lagged effect (i.e., probability of flowering) and its potential environmental drivers [@metcalfStatisticalModellingAnnual2015]. These coupled data sets are rare [sensu @eversLaggedDormantSeason2021], in part because studies to disentangle lagged effects can be challenging to design and maintain [@kuss_evolutionary_2008]. Second, the methods for identifying lagged effects and modeling their demographic impacts can be challenging to implement. Many of the statistical methods have stringent data requirements [@metcalfStatisticalModellingAnnual2015] and assumptions, while the including complex biological processes in demographic models can render them less tractable. Addressing these obstacles is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.\

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellner_integral_2006; @rees_building_2014; @rees_integral_2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguerogomez_compadre_2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morris_quantitative_2002 and @crone_how_2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooks_statistical_2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @rees_evolving_2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.\

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) for all IPMs the projections of $\lambda$ forest fragments would be lower than for continuous forest, but that the difference would be proportionately greater for IPM with lagged effects.\

@@

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kress_diversity_1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H. acuminata* - grow primarily in the forest understory [@kress_self-incompatibility_1983; @ribeiro_influence_2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@bruna_heliconia_2004; @stouffer_forest_1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaard_lessons_2001.\

A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments. All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H. acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999]. On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

We projected the growth rate and structure of *Heliconia acuminata* populations in Continuous Forest and Forest Fragments with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models.\

@@

Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent. IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].\

@@

We used the annual census data [@bruna_are_2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' [@bruna_are_2003]. For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic. To model size-specific fecundity we used data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots (to avoid altering within-plot recruitment) and experimentally-derived estimates of seed germination and seedling establishment [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\

**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@bruna_demography_2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\

**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. The kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\

@@

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, but discarded the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and then back-transformed to allow for direct comparison with projections of $\lambda$ from deterministic models. The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models. Finally, we estimated the 95% confidence intervals for each IPMs projections of $\lambda$. To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector. We then re-fit the vital rate models for growth, survival, and flowering for each of these bootstrapped population and constructed new IPMs for each population as above (the models for germination and establishment rate, fruits per flowering plant, and seeds per fruit were not refit because these were estimated using different data sets). The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manly2018randomization; @caswell2001]. This workflow was managed using the `targets` R package [@targets], which also allowed us to track computational time spent processing and analyzing each class of IPM.\

@@

##### Comparison of CF vs FF

-   For det need to bootstrap

-   for within stoch and lag: glm/t-test.

-   Comparison of pop structure in figure (test for increasing variance with variance / F or permutation test). Villase∼nor, J. A., & González-Estrada, E. (2024). A nonparametric test for homogeneity of variances. Communications in Statistics - Theory and Methods, 54(2), 646–654. https://doi.org/10.1080/03610926.2024.2316274 OR https://www.datanovia.com/en/lessons/homogeneity-of-variance-test-in-r/#google_vignette

@@

det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
det_time<-as.numeric(det_time[1,2])
sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
sker_time<-as.numeric(sker_time[1,2])
spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
spar_time<-as.numeric(spar_time[1,2])
```

### Lambda Comparison

>>>>

---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
        <!-- \usepackage[utf8]{inputenc} -->
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}
        \usepackage{caption}
        \captionsetup[figure]{labelfont={bf},name={fig.},labelsep=period}
        \raggedright
        \usepackage{setspace}\doublespacing
        \setlength\parindent{22pt}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
# library(vartest)
library(purrr)
library(broom)
```

```{r}
#load targets
withr::with_dir(here(), {
  tar_load(c(
    aic_tbl_df,
    lambda_table,
    fig_pop_states
  ))
})

@@

@@

<!-- There is increasing evidence for delays of months or even years between current environmental conditions and changes in the growth, survival, or reproduction of long-lived organisms.  -->

@@

Current environmental conditions can have large and immediate effects on the growth, survival, or reproduction of long-lived organisms. There is also mounting evidence, however, for intervals that range from months to years between environmental conditions and the resulting changes in demographic vital rates. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckermanPopulationDynamicConsequences2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one at a later stage of the reproductive process. The delay or intensity of lagged effects can also be location- or habitat-specific, with individuals in some sites or habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.

Because Lagged Effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics [@beckermanPopulationDynamicConsequences2002]. While recent studies suggest this could indeed the case [e.g., @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015], broader efforts to test this hypothesis have face a number of obstacles [@metcalfStatisticalModellingAnnual2015]. Empirical tests of putative lagged effects because these are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008a]. While one could use conceivably use observational data to detect lagged effects, doing so requires long-term data on both the putative lagged effect (e.g., reduced probability of flowering) and its potential environmental drivers (e.g., drought during early floral development), and such coupled data sets are rare [sensu @eversLaggedDormantSeason2021]. In addition, the methods for both identifying lagged effects and modeling their demographic impacts can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015], while the including complex biological processes in demographic models can render them less tractable or amenable to analysis using currently available tools. Addressing these challenges is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002 and @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooksStatisticalModelingPatterns2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.

@@

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H. acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\

A complete description of our demographic methods, data, and analyses to date can be found in @brunaDemographyUnderstoryHerb2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments (i.e., CF and FF, respectively). All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants. These surveys were complemented by data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots to avoid altering within-plot recruitment. We also conducted experiments to quantify the probability of seed germination and seedling establishment in both forest fragments and continuous forest [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].

*Heliconia acuminata* in our site begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. On average plots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).

@@

We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs - Deterministic, Stochastic, and Stochastic with Lagged Environmental Effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models. Our modeling workflows were managed using the `targets` R package [@targets]; in adddition to ensuring reproducibility this allowed us to track computational time spent processing and analyzing each class of IPM.\

@@

Note that in @eq-R both the probability that new seedlings survive their first year, $s_s$, and their size at the end of this year, $G(z^\prime, z)$, are size-independent. IPMs can include transitions between individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'mature plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treat seedlings as a distinct and discrete category because they have lower survival and growth in their first year than comparably sized plants [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2022].\

@@

We used the annual census data [@brunaArePlantPopulations2003] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (i.e., Fragments, Continuous Forest; the data from all plots within a habitat class were pooled to create a single 'summary population' for the CF and FF habitats [@brunaArePlantPopulations2003]. For established plants these were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` library [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were Leptokurtic. To model size-specific fecundity we used the data on fruits per flowering plant [@brunaLeafNumberLeaf2021], seeds per fruit [@brunaDatasetHeliconiaAcuminata2014], and the experimentally-derived estimates of seed germination and seedling establishment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. <!--# Emilio: add more info here -->\

**(2) Stochastic IPMs:** To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@brunaDemographyUnderstoryHerb2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswell2001].\

**(3) Stochastic IPMs with lagged effects of precipitation on vital rates:** We explicit modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2022. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2022]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. Though the kernels of successive iterations are not entirely independent -- the SPEI values used to calculate vital rates include values used in the previous two iterations -- they are ergodic.\

@@

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels<!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, but discarded the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and then back-transformed to allow for direct comparison with projections of $\lambda$ from deterministic models. The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models.

Finally, We estimated the 95% confidence intervals for each IPMs projections of $\lambda$. To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector. We then re-fit the vital rate models for growth, survival, and flowering for each of these bootstrapped population and constructed new IPMs for each population as above (the models for germination and establishment rate, fruits per flowering plant, and seeds per fruit were not refit because these were estimated using different data sets). The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manly2018randomization; @caswell2001].

@@

##### $\lambda$ in Continuous Forest vs. Forest Fragments.

To determine if the deterministic projections of $\lambda$ for Continuous Forest and Forest Fragments were significantly different between habitats we used the randomization test procedure described by @caswell2001. Briefly, we randomly assigned plants with their demographic history among two populations, $R_{1}$ and $R_{2}$, that were equal in size to the original CF and FF populations. We then calculated $\lambda^{R_1}$, $\lambda^{R_2}$, and the absolute value of the difference between the two (i.e., $\theta$). This was repeated N = 1000 times, after which we determined the proportion of simulations in which $\theta$ was greater than the observed difference between $\lambda^{CF}$ and $\lambda^{FF}$ (i.e., $P[\theta \geq \theta_{obs}]$).

<!-- -   for within stoch and lag: glm/t-test. -->

<!-- see https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577 for suggestions on text ffor describing and reporting results -->

We used a Generalized Linear Model to compare the projections of $\lambda$ from the two stochastic IPMs. We modelled the lambda as a function of IPM type (kernel-resampling vs parameter-resampling) and Habitat (continuous forest vs. fragments), and the interaction of IPM type and Habitat. Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity function in our model. We verified the model assumptions by plotting residuals versus fitted values.\

```{r glm-lambda}

path <- here("docs/figures/stoch_lambdas.csv")
stoch_lambdas<-read_csv(path)
lambda_glm <- glm(lambda ~ habitat*ipm_type,
             family = gaussian(link = "identity"), data = stoch_lambdas)

# summary(lambda_glm)
# (over)dispersaion
# pearson's residuals
# sum(resid(lambda_glm, type="pearson")^2)/lambda_glm$df.resid
#deviance
# lambda_glm$deviance/model$df.residual
# plot(lambda_glm, which = 1)
# plot(lambda_glm, which = 2)
# https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577
# acf(residuals(lambda_glm)) ## check autocorrelation
```

##### Population Structure.

Finally, we compared the structure of populations through the first 250 time steps in each habitat by each of the stochastic IPMs (i.e, kernel- vs. parameter-resampling). At each time step, we assigned the individuals to one of four stage classes based on plant size, probability of survival, and probability of reproduction (see Table @tbl-plantcats). We then calculated the proportion of the population that was in each stage class in each habitat for each IPM type. Comparing the means and variances of these proportions will allow us to determine (1) if the two IPMs project different population structures over time, (2) if the population structure for a given IPM is consistent across habitat type. Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination. These analyses were conducted using the R libraries `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].

@@

det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
det_time<-as.numeric(det_time[1,2])
sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
sker_time<-as.numeric(sker_time[1,2])
spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
spar_time<-as.numeric(spar_time[1,2])

path_det_stats <- here("docs","figures","det_randomization_test.csv")
det_stats<-read_csv(path_det_stats) %>% 
  mutate(p_val=100-perc_below)
```
